### 2025-06-11 chapters/methodology.tex
updated sections 3.3, 3.4 & references (still in progress)

To get a broader picture of system performance—particularly one that includes moderate memory use and more complex workloads—the CoreMark benchmark was also used. Developed by the Embedded Microprocessor Benchmark Consortium (EEMBC), CoreMark was designed to overcome some of the limitations of Dhrystone by providing a workload that’s a bit closer to real-world embedded applications. It includes tasks like linked list manipulation, matrix operations, and finite state machine processing, all of which are common in software running on embedded systems. While CoreMark is still CPU-focused, it introduces a moderate level of memory interaction through dynamic data structures, which allows it to reveal performance issues that Dhrystone might miss—especially those related to cache usage or memory protection features within a TEE.

The version of CoreMark used in this study came from the SiFive repository, which provides a version optimized for RISC-V systems and ready to work with Keystone. Like Dhrystone, CoreMark was run in both native and enclave modes to compare performance under each. Its results are reported in terms of "Iterations Per Second" (IPS), which tells us how many complete runs of the benchmark are executed every second. Since CoreMark includes both arithmetic and memory operations, IPS gives a more holistic view of how the system is performing—not just the processor, but also the memory subsystem.

@@

\section{Parameter Variation Strategy}

\begin{comment}
In order to conduct a comprehensive and methodologically sound performance evaluation of the Keystone Trusted Execution Environment (TEE), a structured parameter variation strategy was employed. The goal of this strategy was to explore how different system and workload configurations affect performance, both inside and outside of the enclave context. By systematically varying a set of key parameters, it becomes possible to isolate the sources of performance overhead and assess the sensitivity of the system to specific operational conditions.

A fundamental aspect of the evaluation involved comparing benchmark execution in two distinct contexts: native execution versus enclave execution. In the native configuration, benchmarks were executed directly within the emulated Linux environment running on QEMU, without invoking Keystone’s enclave infrastructure. This configuration provided a performance baseline that reflects the raw computational capability of the underlying system without the influence of TEE-specific isolation mechanisms. In contrast, enclave execution involved running the same benchmark binaries within Keystone-enforced secure containers. This allowed for a direct measurement of the overhead introduced by features such as memory isolation, limited system call interfaces, and the additional control flow transitions required by the Security Monitor.

To ensure statistical significance and reproducibility, each benchmark was executed multiple times under identical conditions. Specifically, each configuration (e.g., native or enclave) was run ten times, and the results were aggregated to compute average execution time, throughput (expressed as DMIPS for Dhrystone or IPS for CoreMark), as well as the standard deviation across runs. This repetition helped mitigate the impact of transient system noise or timing anomalies that can occur in virtualized environments, especially under QEMU emulation. Additionally, the inclusion of standard deviation in the reported results provided insights into the consistency and stability of the enclave performance.

Another important axis of variation was workload size. For the CoreMark benchmark, which allows for configurable execution parameters, experiments were conducted using different iteration counts and internal data structure sizes. This enabled the evaluation of how enclave performance scales with increasing computational intensity and memory utilization. Because enclave memory is subject to isolation and access restrictions, increasing the memory footprint can expose performance bottlenecks that would not be apparent under smaller workloads. While Dhrystone does not support direct workload scaling through runtime arguments, multiple versions of the benchmark were compiled with different loop unroll factors to simulate variation in workload complexity. Although less precise than CoreMark’s controls, this approach still provided a meaningful indication of how performance trends shift with increasing instruction throughput requirements.

The influence of system-level configuration options was also examined. Keystone offers various compile-time and runtime flags that affect behavior and performance, most notably the debug mode. When enabled, debug mode activates additional runtime checks, logging mechanisms, and diagnostic outputs intended to aid developers during enclave development and testing. These features, while useful during implementation, introduce measurable overhead. By executing benchmarks with and without debug mode enabled, the experiments were able to quantify the cost of enhanced observability and determine whether such features significantly degrade performance.

Memory allocation parameters were also varied to understand their role in shaping enclave performance. In particular, the heap and stack sizes assigned to benchmarks were adjusted to evaluate the sensitivity of the system to memory layout and availability. This was particularly relevant for CoreMark, which dynamically allocates and manipulates internal data structures. Larger memory allocations could potentially lead to increased cache pressure or page faults within the enclave, highlighting the importance of efficient memory management in TEE contexts.

Measurement accuracy was further reinforced by cross-validating timing results. Execution time was captured using both the benchmarks’ internal timing functions and QEMU’s emulated timer infrastructure. This dual-timing approach ensured that timing results were not distorted by emulator artifacts or low-resolution timers. In cases where discrepancies were observed, additional investigation was conducted to determine the root cause, and only consistent, corroborated results were included in the final analysis.

Finally, to explore the potential impact of system-level contention, a set of exploratory experiments introduced artificial background load during native execution. Although Keystone enclaves currently operate in a single-threaded, non-preemptive mode, testing under load offered preliminary insights into how enclave performance might be influenced in future multi-threaded or multi-tenant scenarios, especially in shared virtualized environments. While these tests were not a primary focus of the evaluation, they provided valuable context for interpreting variability in the results.
\end{comment}

\section{Data Collection and Analysis Procedures}

>>>>

To get a broader picture of system performance—particularly one that includes moderate memory use and more complex workloads—the CoreMark benchmark \cite{gal2012exploring} was also used. Developed by the Embedded Microprocessor Benchmark Consortium (EEMBC), CoreMark \cite{gal2012exploring} was designed to overcome some of the limitations of Dhrystone by providing a workload that’s a bit closer to real-world embedded applications. It includes tasks like linked list manipulation, matrix operations, and finite state machine processing, all of which are common in software running on embedded systems. While CoreMark \cite{gal2012exploring} is still CPU-focused, it introduces a moderate level of memory interaction through dynamic data structures, which allows it to reveal performance issues that Dhrystone might miss—especially those related to cache usage or memory protection features within a TEE.

The version of CoreMark \cite{gal2012exploring} used in this study came from the SiFive repository available on GitHub at \url{https://github.com/sifive/benchmark-coremark}, which provides a version optimized for RISC-V systems and ready to work with Keystone. Like Dhrystone \cite{weiss2002dhrystone}, CoreMark \cite{gal2012exploring} was run in both native and enclave modes to compare performance under each. Its results are reported in terms of "Iterations Per Second" (IPS), which tells us how many complete runs of the benchmark are executed every second. Since CoreMark includes both arithmetic and memory operations, IPS gives a more holistic view of how the system is performing—not just the processor, but also the memory subsystem.

@@

\section{Parameter Variation Strategy}
\label{sec:param-variation}

To examine the influence of microarchitectural cache parameters on enclave performance, this study employed a structured variation of Level-1 (L1) cache configurations using QEMU's cache modeling capabilities. The experimental objective was to isolate the impact of cache size, associativity, eviction policy, and core count on the performance of secure and non-secure workloads in a controlled emulation environment.

Cache modeling was enabled through the QEMU TCG plugin framework \cite{mandour2021cache}, which simulates idealized, per-core instruction and data caches. While not microarchitecturally cycle-accurate, this plugin supports a configurable cache hierarchy, enabling systematic exploration of cache behaviors relevant to trusted execution environments.

Four key experimental scenarios were defined:

\begin{enumerate}
    \item \textbf{Baseline Configuration:} A minimal cache setup with 8\,KB instruction and data caches, 4-way associativity, and LRU eviction.
    \item \textbf{Larger Cache Scenario:} An expanded L1 cache configuration with 32\,KB size and 8-way associativity to evaluate the effect of reduced cache miss rates.
    \item \textbf{Multicore Simulation:} A four-core configuration with private per-core caches to study inter-core effects such as contention and cache coherence approximations.
    \item \textbf{Eviction Policy Sensitivity:} A fixed-size cache using alternative eviction strategies (FIFO and random) to assess their influence on enclave memory behavior.
\end{enumerate}

All scenarios maintained a fixed block size of 64 bytes to preserve spatial locality characteristics. The same set of benchmarks—Dhrystone and CoreMark—was executed under each configuration, both in native and enclave modes, to quantify differences in execution time, throughput, and cache performance.

The plugin-generated metrics included instruction and data cache accesses, miss rates, and identification of cache-thrashing instructions. While detailed configuration parameters, plugin invocation commands, and sample output logs are essential for reproduction, they are provided in Appendix~\ref{appendix:cache-plugin} to avoid interrupting the main methodological flow.

By varying one parameter set at a time while controlling for others, this approach facilitated a clear attribution of observed performance differences to specific architectural characteristics. The resulting data informed the broader analysis of how cache-aware design considerations can impact enclave performance in systems based on the Keystone framework.

\section{Data Collection and Analysis Procedures}
\label{sec:data-collection}

To ensure that the performance evaluation of enclave execution is both statistically robust and reproducible, a systematic approach was taken to data collection and analysis. All benchmarks were executed within the same virtualized RISC-V environment described in Section~\ref{chap:methodology}, with controlled cache parameters as outlined in Section~\ref{sec:param-variation}.

Each benchmark (Dhrystone and CoreMark) was executed under two configurations: native execution and enclave execution. For every configuration, a minimum of ten independent iterations was conducted to account for runtime variability. Execution time was measured using wall-clock timing as reported by the host Linux environment inside QEMU. Benchmark-specific performance metrics—Dhrystone MIPS (DMIPS) and CoreMark iterations per second (IPS)—were recorded after each run.

The collected data for each experiment included:

\begin{itemize}
    \item Total execution time (seconds)
    \item Benchmark throughput (DMIPS or IPS)
    \item Standard deviation and coefficient of variation
    \item Cache miss rates (instruction/data) when applicable
\end{itemize}

To mitigate the effects of transient system behavior, the first run of each benchmark was discarded as a warm-up iteration. The remaining samples were then averaged, and both the mean and standard deviation were computed. These values were used to assess not only the central tendency of performance, but also the stability of enclave execution under different cache configurations.

For cache-sensitive experiments, data was further enriched using QEMU's cache modeling plugin \cite{mandour2021cache}. This provided detailed logging of instruction and data accesses, cache miss counts, and identification of high-miss instructions. From these logs, miss rates were extracted and correlated with execution behavior. Selected logs and plugin outputs are included in Appendix~\ref{appendix:cache-plugin}.

Statistical analysis was conducted using standard descriptive statistics. Where relevant, normalized overhead values were computed to express enclave slowdown as a percentage increase in execution time or a reduction in throughput compared to the native baseline. These comparative metrics provide a clearer view of the cost of enclave isolation across various architectural conditions.

The consistent use of controlled environments, repeated measurements, and structured analysis ensures that the results presented in subsequent chapters are both methodologically sound and analytically valid.
